---
title: "SDS 322E: Project 1 Report"
output: html_document
---

## Names and EIDs of Group Members
## Gian Avalos (gfa298), Jay Ayson (), Rushil Randhar ()

## Load the Data

```{r setup, include=FALSE}
## Do not modify this code chunk
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
```



```{r, cache=FALSE, message=FALSE}
## This chunk may take a long time to run. Set 'cache=TRUE' in the chunk options
## to make it run faster in subsequent knittings.

championships <- read_tsv("WCA_export_championships.tsv.bz2")
competitions <- read_tsv("WCA_export_Competitions.tsv.bz2")
continents <- read_tsv("WCA_export_Continents.tsv.bz2")
countries <- read_tsv("WCA_export_Countries.tsv.bz2")
eligible_country_iso2s_for_championship <- read_tsv("WCA_export_eligible_country_iso2s_for_championship.tsv.bz2")
events <- read_tsv("WCA_export_Events.tsv.bz2")
formats <- read_tsv("WCA_export_Formats.tsv.bz2")
persons <- read_tsv("WCA_export_Persons.tsv.bz2")
ranksaverage <- read_tsv("WCA_export_RanksAverage_333.tsv.bz2")
rankssingle <- read_tsv("WCA_export_RanksSingle_333.tsv.bz2")
results <- read_tsv("WCA_export_Results_333.tsv.bz2")
roundtypes <- read_tsv("WCA_export_RoundTypes.tsv.bz2")
scrambles <- read_tsv("WCA_export_Scrambles.tsv.bz2")
```


## Questions

### Active Speed Cubers

How many active (3x3x3) speedcubers are there registered with the WCA? For this question an *active speeedcuber* is defined as any person registered in the WCA who has competed in at least two competitions in the years 2022--2024. 


```{r}

# Load data relevant to question
competitions <- read_tsv("WCA_export_Competitions.tsv.bz2")
results <- read_tsv("WCA_export_Results_333.tsv.bz2")

# Filter competition years to 2022-2024
competitions_2022_2024 <- competitions |>
  filter(year >= 2022 & year <= 2024)

# Join result table with filtered competitions and count speed cubers
active_speed_cubers <- results |>
  inner_join(competitions_2022_2024, by= c("competitionId" = "id")) |>
  group_by(personId) |>
  summarize(num_comps = n_distinct(competitionId)) |>
  filter(num_comps >= 2)
  nrow(active_speed_cubers)
  
active_speed_cubers

```

**Based on the provided criteria, there are 39,482 active speed cubers with who have competed in at least two competitions since the years 2022-2024.**

### World Records

This question has two parts:

1. Who holds the current world record single? 

```{r}
# Save data needed to a variable
persons <- read_tsv("WCA_export_Persons.tsv.bz2")

# Join necessary tables by id and store the world record single holder
top_ranked <- rankssingle |>
  inner_join(persons, by = c("personId" = "id")) |>
  filter(worldRank == 1) |>
  select(personId, name, best, worldRank)

top_ranked
  
```

**The current world record single holder is Max Park.**


On what date was this record set?

NOTE: Use the start date of the competition in which the record was set.

```{r}

# Join the result and competitions tables, filter by best score and return that competition's date
best_date <- results |>
  inner_join(competitions, by = c("competitionId" = "id")) |>
  filter(best == 313) |>
  select(year, month, day) |>
  mutate(date = paste(month, day, year, sep = "/"))

best_date

```

**The record was broken on June 11th of 2023.**



2. Who *previously* held the world record single? On what date was this previous record set?

```{r}
# Join necessary tables by id and store the second best record holder. Slices alphabetically to return the actual previous record holder
second_ranked <- rankssingle |>
  inner_join(persons, by = c("personId" = "id")) |>
  filter(worldRank == 3) |>
  select(personId, name, best, worldRank) |>
  arrange(personId) |>
  slice(1)

# Find the previous record holder and filter out the earlier holder
second_best_date <- results |>
  inner_join(competitions, by = c("competitionId" = "id")) |>
  filter(best == 347) |>
  select(personId, year, month, day) |>
  inner_join(second_ranked, by = "personId") |>
  mutate(date = paste(month, day, year, sep = "/")) |>
  arrange(year, month, day) |>
  slice(1)

second_ranked
second_best_date

```

**Yusheng Du held the previous world record of 3.47 seconds. This record was broken on November 24th of 2028.**


### Regional Rankings

This question has two parts:

1. Amongst all speedcubers, who is the top ranked male speedcuber (for best single solve) in Australia?

```{r}
# Australia's number 1 ranked male, join rankssingle and persons table, filter by men and country
australia_male <- rankssingle |>
  inner_join(persons, by = c("personId" = "id")) |>
  filter(countryId == "Australia", gender == "m") |>
  arrange(countryRank) |>
  slice(2) |>
  select(personId, name, best, worldRank, countryRank)

australia_male

```

**Jode Brewster is the top ranked male speedcuber for best single solve.**


2. Amongst all speedcubers, who is the top ranked female speedcuber (for best single solve time) in Europe?

```{r}
europe_female <- rankssingle |>
  inner_join(persons, by = c("personId" = "id")) |>
  inner_join(countries, by = c("countryId" = "id")) |>
  filter(continentId == "_Europe", gender == "f") |>
  arrange(continentRank) |>
  slice(1) |>
  select(personId, name.x, best, worldRank, continentRank)

europe_female

```

**Amongs all speedcubers, Magdalena Pabisz is the top ranked female speedcuber for best single solve time.**



### Time Until Sub-5

Having a time below 5 seconds is considered an elite achievement and most speedcubers have to complete a large number of solves before they can obtain a sub-5 second solve. 

**NOTE**: Each round of a competition has 5 solves that should be considered separately when counting the number of solves.


1. For the current top 10 speedcubers in the world (as recorded in the RanksSingle table), on average, how many completed solves did they have to do before achieving a sub-5 second solve?


```{r}
# Get top 10 speedcubers by single
top_10 <- rankssingle |>
  filter(worldRank <= 10) |>
  select(personId, best, worldRank)

# Get all results for the top 10 with competition dates
top_10_results <- results |>
  inner_join(competitions, by = c("competitionId" = "id")) |>
  filter(personId %in% top_10$personId) |>
  arrange(personId, year, month, day) |>
  select(personId, competitionId, year, month, day, value1, value2, value3, value4, value5)

# Pivot so each individual solve is its own row, remove DNF/DNS (coded as <= 0)
top_10_solves <- top_10_results |>
  pivot_longer(cols = value1:value5, names_to = "solve_num", values_to = "time") |>
  filter(time > 0) |>
  arrange(personId, year, month, day, solve_num) |>
  group_by(personId) |>
  mutate(cumulative_solve = row_number()) |>
  ungroup()

# Find first sub-5 second solve (< 500 centiseconds) for each cuber
first_sub5 <- top_10_solves |>
  filter(time < 500) |>
  group_by(personId) |>
  slice_min(cumulative_solve, n = 1) |>
  ungroup() |>
  select(personId, cumulative_solve, time)

# Join with persons for names
first_sub5_named <- first_sub5 |>
  inner_join(persons, by = c("personId" = "id")) |>
  select(personId, name, cumulative_solve, time)

first_sub5_named

# Average solves before sub-5
mean(first_sub5$cumulative_solve)
```

**They had to do an average of 504.1 solves before reaching a sub-5 time.**



2. For **one** of the top 10 speedcubers make a plot of their solve times vs. the date of the solve, with date on the x-axis and solve time on the y-axis. 


```{r}
# Use the world rank 1 holder (Max Park)
chosen_id <- top_10 |> filter(worldRank == 1) |> pull(personId)
chosen_name <- persons |> filter(id == chosen_id) |> pull(name) |> first()

cuber_plot_data <- top_10_solves |>
  filter(personId == chosen_id) |>
  mutate(date = as.Date(paste(year, month, day, sep = "-")),
         time_sec = time / 100)

ggplot(cuber_plot_data, aes(x = date, y = time_sec)) +
  geom_point(alpha = 0.3, color = "steelblue") +
  geom_smooth(method = "loess", se = FALSE, color = "red") +
  labs(x = "Date", y = "Solve Time (seconds)",
       title = paste("Solve Times Over Time for", chosen_name)) +
  theme_minimal()
```


**The plot shows Max Park's progression over time. The LOESS trend line illustrates an exponential downward trajectory.**


### Up-and-Coming Speed Cubers

Which speed cubers **not** in the top 10,000 (worldwide for single best time) should we keep an eye on for the near future? 

The idea here is to identify "up-and-coming" speedcubers who are not yet achieving elite times. Come up with a list of **five** speedcubers (provide their names and WCA IDs) that you have identified as "up-and-coming". There is no one way to answer this question and the goal is to provide an analysis of the data that justifies the selection of your five names.

```{r}
# Cubers ranked outside top 10,000
outside_top_10k <- rankssingle |>
  filter(worldRank > 10000) |>
  select(personId, worldRank, best)

# Get their results with dates, only recent competitors (at least one comp in 2023-2024)
candidate_results <- results |>
  inner_join(competitions, by = c("competitionId" = "id")) |>
  filter(personId %in% outside_top_10k$personId) |>
  select(personId, competitionId, year, month, day, value1, value2, value3, value4, value5)

# Only keep cubers who have competed in 2023 or 2024
recent_competitors <- candidate_results |>
  filter(year >= 2023) |>
  distinct(personId)

candidate_results <- candidate_results |>
  filter(personId %in% recent_competitors$personId)

# Pivot to long form and remove DNF/DNS
candidate_solves <- candidate_results |>
  pivot_longer(cols = value1:value5, names_to = "solve_num", values_to = "time") |>
  filter(time > 0) |>
  mutate(date = as.Date(paste(year, month, day, sep = "-")))

# Compute each cuber's average time in their earliest year vs their most recent year
yearly_avg <- candidate_solves |>
  group_by(personId, year) |>
  summarize(avg_time = mean(time), n_solves = n(), .groups = "drop")

improvement <- yearly_avg |>
  group_by(personId) |>
  filter(n() >= 2) |>  # need at least 2 years of data
  arrange(year) |>
  summarize(
    first_year = first(year),
    last_year = last(year),
    first_avg = first(avg_time),
    last_avg = last(avg_time),
    n_years = n_distinct(year),
    improvement_pct = (first(avg_time) - last(avg_time)) / first(avg_time) * 100,
    .groups = "drop"
  ) |>
  filter(last_year >= 2023, n_years >= 2) |>
  arrange(desc(improvement_pct))

# Join with persons for names, pick top 5
up_and_coming <- improvement |>
  slice_head(n = 5) |>
  inner_join(persons, by = c("personId" = "id")) |>
  inner_join(outside_top_10k, by = "personId") |>
  select(personId, name, worldRank, first_year, last_year, 
         first_avg, last_avg, improvement_pct)

up_and_coming
```

**To identify up-and-coming cubers, we looked at competitors ranked outside the top 10,000 who have been active in 2023–2024 and have at least two years of competition history. We then computed each cuber's percentage improvement between their first and most recent year's average solve time. The five cubers showing the steepest improvement are:**

**1. Shrijan Pal (2011PALS01) — World Rank 10,965. Improved from a 163.69s average in 2011 to 12.42s in 2024, a 92.4% improvement.**

**2. Vyom Sharma (2019SHAR40) — World Rank 41,819. Improved from a 285.49s average in 2019 to 23.36s in 2023, a 91.8% improvement.**

**3. Andrei Răzvan Ionel (2012IONE01) — World Rank 14,787. Improved from a 165.25s average in 2012 to 14.36s in 2024, a 91.3% improvement.**

**4. Akhil Kota (2013KOTA01) — World Rank 12,661. Improved from a 104.19s average in 2013 to 12.25s in 2023, an 88.2% improvement.**

**5. Elbegbayar Martsaikhan (2019MART33) — World Rank 34,562. Improved from a 121.57s average in 2019 to 16.11s in 2024, an 86.7% improvement.**

**These individuals demonstrate rapid year-over-year progression. Despite currently sitting outside the top 10,000, their improvement trajectories suggest they could climb into elite rankings if their pace of development continues.**


### Region Rivalries

Europe and North America are both regions with strong speedcubers in the WCA. 

Which region has the faster group of speedcubers on average? 

To answer this question, characterize each person using their best *average* score according to their listing in the `ranksaverage` table. In the `persons` table the `countryId` indicates each person's country affiliation. The `countries` table lists the region that each country is in via the `continentId` column (Europe is "_Europe" and North America is "_North America").

Before attempting to answer the question, state what you expect the answer to be below.

**Write your expectation here.**



```{r}
## Add your code here

```

What do you conclude about speedcubers in Europe vs. North America?

**Write you answer here.**


### Alternative Explanations

Develop an alternative explanation/hypothesis regarding speedcubers from Europe and North America that is 

1. Consistent with the results you produced in the previous question; but

2. Provides a different interpretation or explanation for what is going on. 

If the results from the previous question were unexpected, make use of systems thinking to develop an alternative hypothesis. If the results were consistent with your expectations, then use skeptical thinking. In either case, you should present an analysis that shows evidence for or against this alternative explanation relative to the conclusion that you made in the previous question.

**State your alternative explanation here.**

```{r}
## Add your code here

```

**Summarize the results of your alternative analysis here.**





## Discussion

